%!TEX root = ../main.tex

\chapter*{Publications}

Chapters \ref{ch:2} and \ref{ch:4} are respectively based on the following journal articles:

\begin{itemize}
\item de Fleurian, R., \& Pearce, M. T. (2021). Chills in music: A systematic review. \emph{Psychological Bulletin}, \emph{147}(9), 890--920. \url{https://doi.org/10.1037/bul0000341}
\item de Fleurian, R., \& Pearce, M. T. (2021). The relationship between valence and chills in music: A corpus analysis. \emph{i-Perception}, \emph{12}(4), 1--11. \url{https://doi.org/10.1177/20416695211024680}
\end{itemize}

\noindent Chapters \ref{ch:3} and \ref{ch:5} are currently being adapted for the following submissions:

\begin{itemize}
\item de Fleurian, R., \& Pearce, M. T. (2022). \emph{Longitudinal study of chills in music: Effects of musical content, stylistic preference, and familiarity}. Manuscript in preparation.
\item de Fleurian, R., Benetos, E., Clemente, A., \& Pearce, M. T. (2022). \emph{Automatic detection of chills in music}. Manuscript in preparation.
\end{itemize}

\noindent The data collected in Chapters \ref{ch:2} and \ref{ch:5} was released in two publicly available datasets:

\begin{itemize}
\item de Fleurian, R., \& Pearce, M. T. (2021). \emph{Chills in Music (ChiM)}. OSF. \url{https://doi.org/10.17605/osf.io/uyg7m}
\item de Fleurian, R., \& Pearce, M. T. (2021). \emph{Onsets of Chills in Music (oChiM)}. OSF. \url{https://doi.org/10.17605/osf.io/x59fm}
\end{itemize}

\noindent Early versions of the work presented in this thesis appeared in the following conference proceedings and programmes:

\begin{itemize}
\item de Fleurian, R., \& Pearce, M. T. (2019). Musical chills: Effects of stimulus properties, stylistic preference and familiarity. In P. Martens, \& F. Upham (Eds.), \emph{Abstracts of the 2019 biennial meeting of the Society for Music Perception and Cognition, SMPC 2019} (p. 37). 
\item de Fleurian, R., \& Pearce, M. T. (2019). Effects of stimulus properties, stylistic preference and familiarity on musical chills. In \emph{Programme of the SEMPRE Graduate Conference 2019} (p. 26). 
\item de Fleurian, R., \& Pearce, M. T. (2018). Musical chills: Stimulus properties, stylistic preference and familiarity. In P. Kudumakis, \& S. Dixon (Eds.), \emph{Proceedings of the Digital Music Research Network workshop, DMRN+13} (p. 19). 
\end{itemize}

\noindent Finally, the following article was also produced during the doctoral study period:

\begin{itemize}
\item de Fleurian, R., Harrison, P. M. C., Pearce, M. T., \& Quiroga-Martinez, D. R. (2019). Reward prediction tells us less than expected about musical pleasure. \emph{Proceedings of the National Academy of Sciences of the United States of America}, \emph{116}(42), 20813--20814. \url{https://doi.org/10.1073/pnas.1913244116}
\end{itemize}
